Evaluating FAIR Digital Object as a distributed object system

### 2023-01-31 latex/main.tex
some citation keys fixed

The FAIR principles \cite{wilkinson_fair_2016} and \cite{working_group_doip_2022} and \cite{noauthor_info_nodate} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures. They have been widely recognised by the research community and policy makers as a goal to strive for. In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \cite{u8Cy0psL}. The EOSC Interoperability Framework \cite{aCye3KpE} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.

Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object \cite{IHLT6hye} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures. The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.

In other discourse, Linked Data \cite{19s3yyxDn} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \cite{EJDjHrUv,GJHYN6xW}. Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{jHwttlTs}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \cite{99MS5xE0,J2H8yssV}). While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \cite{DNTCUjpC,Pa67pUtR}.

One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality \cite{uxWjmlio,v0WwFHgT} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{5V3GqSzR}.

>>>>

The FAIR principles \cite{wilkinsonFAIRGuidingPrinciples2016e} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures. They have been widely recognised by the research community and policy makers as a goal to strive for. In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \cite{monsCloudyIncreasinglyFAIR2017b}. The EOSC Interoperability Framework \cite{corchoEOSCInteroperabilityFramework2021b} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.

Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object \cite{schultesFAIRPrinciplesDigital2019a} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures. The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.

In other discourse, Linked Data \cite{bizerLinkedDataStory2009a} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \cite{boninodasilvasantosFAIRDataPoints2016a,hasnainAssessingFAIRData2018a}. Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{verborghSemanticWebIdentity2020a}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \cite{merono-penuelaConclusionFutureChallenges2021a,neumannAnalysisPublicREST2021a}). While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \cite{turcoaneLinkedDataJSONLD2014a,wilkinsonWorkflowsWhenParts2022b}.

One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality \cite{stefiDevelopersMakeUnbiased2015,stefiDevelopReuseTwo2015a} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{merono-penuelaWebDataApis2021b}.
